Structure constants and short-time simulations of open collective quantum spin systems

### 2019-03-26 soc_squeezing/collective_paper/collective_spin.tex
wording changes and minor typo fixes

In this work, we present a new method to simulate short-time dynamics
of collective spin systems with permutationally-symmetric decoherence.
This method is based on a short-time expansion of exact solutions to
the equations of motion for Heisenberg operators.  Evaluating this
expansion requires using the structure constants of a collective spin
operator algebra; the calculation of these structure constants (in
Appendices \ref{sec:identities}--\ref{sec:prod_general}) is one of the
main technical results of this work, which we hope will enable future
studies to develop new analytical insights into the behavior of
collective spin systems.  We benchmark our method against both exact
and quantum trajectory Monte Carlo calculations of spin squeezing in
accessible parameter regimes, highlighting both advantages and
limitations of the short-time expansion.  Finally, we showcase an
application of our method with calculations that are inaccessible to
other methods.

@@

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$s_{n=\x,\y,\z}\equiv\sigma_n/2$ and
$s_\pm\equiv s_\x\pm is_\y=\sigma_\pm$ with Pauli operators
$\sigma_n$, we denote an operator that acts with $s_n$ on the spin
indexed by $j$ and trivially (i.e.~with the identity $\1$) on all
other spins by $s_n^{(j)}$.  We then define the collective spin
operators $S_n\equiv\sum_{j=1}^Ns_n^{(j)}$ for
$n\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\S_{\v m}}$ as a
basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$, we can expand all
collective spin Hamiltonians in the form
\begin{align}
  H = \sum_{\v m} h_{\v m} \S_{\v m},
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{\v m}^*=h_{\v m^*}$ for
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The time evolution of a particular
Heisenberg operator $\S_{\v m}$ under a Hamiltonian of the form in
Eq.~\eqref{eq:H_general} is then given by
\begin{align}
  \f{d}{dt}\S_{\v m}
  &= i\sum_{\v n}h_{\v n}\sp{\S_{\v n},\S_{\v m}}_-
  + \sum_\J \gamma_\J \D\p{\J;\S_{\v m}} \\
  &= \sum_{\v n} T_{\v m\v n} \S_{\v n},
  \label{eq:EOM}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\J$ is a set of jump operators
with a corresponding decoherence rate $\gamma_\J$;
\begin{align}
  \D\p{\J;\O}
  \equiv \sum_{J\in\J} \p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+}
\end{align}
is the Heisenberg picture dissipator, or Lindblad superoperator, for
all jump operators $J\in\J$; and $T_{\v m\v n}$ is a matrix element of
the time derivative operator.  Decoherence via uncorrelated decay of
individual spins at a rate $\gamma_-$, for example, would be described
by the set of jump operators $\J_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$
with $\gamma_{\J_-}=\gamma_-$.  The commutator in Eq.~\eqref{eq:EOM}
can be determined from the structure constants $f_{\v\ell\v m\v n}$ of
the general product
$\S_{\v\ell}\S_{\v m}=\sum_{\v n}f_{\v\ell\v m\v n}\S_{\v n}$ worked
out in Appendices \ref{sec:identities}--\ref{sec:prod_general}, while
the effects of decoherence from jump operators (i.e.~elements of $\J$)
of the form $\gamma^{(j)}=\sum_n\gamma_ns_n^{(j)}$ and
$\Gamma=\sum_n\Gamma_nS_n$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations one of the main technical contributions of
our work, with potential applications beyond the short-time simulation
method presented here.  These ingredients are sufficient to compute
matrix elements $T_{\v m\v n}$ of the time derivative operator $d/dt$
in Eq.~\eqref{eq:EOM} in most cases of practical interest.

@@

We can expand Heisenberg operators in a Taylor series about $t=0$ to
write
\begin{align}
  \bk{\S_{\v m}}
  = \sum_{k\ge0} \f{t^k}{k!} \bk{\f{d^k}{dt^k} \S_{\v m}}_{t=0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \bk{\S_{\v n}}_{t=0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_1} T_{\v p_1\v p_2} T_{\v p_2\v p_3}
  \cdots T_{\v p_{k-1}\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \bk{\S_{\v m}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \bk{\S_{\v n}}_{t=0}.
  \label{eq:order_truncation}
\end{align}
Unlike the truncation in Eq.~\eqref{eq:weight_truncation}, the nonzero
matrix elements $T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ now tell us
which operators are relevant for computing the expectation value
$\bk{\S_{\v m}}$ to a fixed order $M$.  Note that the relation
$\S_{\v m^*}=\S_{\v m}^\dag$, which by Hermitian conjugation of
Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can be exploited to cut both the
number of expectation values $\bk{\S_{\v m}}_{t=0}$ and matrix
elements $T_{\v m\v n}$ that we may need to explicitly compute roughly
in half.

In principle, the factorial suppression of terms at higher orders of
the Taylor expansion in Eq.~\eqref{eq:time_series} implies for any
given time $t$, there exists a truncation order $M_t$ for which the
truncation error in Eq.~\eqref{eq:order_truncation} is negligibly
small.  In practice, only a maximal truncation order $M_{\t{max}}$ is
accessible with limited computational resources, such that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} only
allows computation of the correlator $\bk{\S_{\v m}}$ to a maximal
time $t_{M_{\t{max}}}$.  As we will see in the following section,
$M_{\t{max}}=35$ will suffice for the accurate computation of
collective spin correlators up to times that are relevant e.g.~for
spin squeezing protocols.

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is the
directionally-unbiased Ramsey squeezing parameter determined by the
maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$ over
that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\v S\equiv\p{S_\x,S_\y,S_\z}$ is a collective spin vector, and
the minimization is performed over all unit vectors $\uv n$ in the
plane orthogonal to the mean spin vector $\bk{\v S}$.  This squeezing
parameter is entirely determined by one- and two-spin correlators of
the form $\bk{S_m}$ and $\bk{S_m S_n}$.  In the case of coherent
dynamics discussed here, these correlators are obtainable via exact
simulations of quantum dynamics in the $\p{N+1}$-dimensional Dicke
manifold of states $\set{\ket{S,m}}$ with net spin $S$ and spin
projection $m$ onto the $z$ axis, i.e.~with
$\bk{S,m|\v S^2|S,m}=S\p{S+1}$ and $\bk{S,m|S_\z|S,m}=m$ for
$m\in\set{-S,-S+1,\cdots,S}$.  In the presence of
permutationally-symmetric decoherence, these correlators are
obtainable with the collective-spin quantum trajectory Monte Carlo
method developed in ref.~\cite{zhang2018montecarlo}.  In this work,
these exact and quantum trajectory simulations will be used to
benchmark the truncated Taylor expansion in
Eq.~\eqref{eq:order_truncation}.

@@

The main lesson from Figure \ref{fig:benchmarking} is that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} gives
essentially exact results right up until a sudden and drastic
departure that is simple to diagnose by inspection.  This drastic
departure occurs because the truncated Taylor expansion neglects
high-weight (i.e.~large-$\abs{\v n}$) operators $\S_{\v n}$ whose
contributions to a correlator of interest $\bk{\S_{\v m}}$ eventually
become non-negligible.  For sufficiently large times $t$, these
individual contributions may generally be large compared to the value
of $\bk{\S_{\v m}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the truncated Taylor expansion and
benchmarking methods when squeezing the initial state $\ket{-\Z}$
rather than $\ket{\X}$: for initial states $\ket{-\Z}$, all
correlators $\bk{\S_{\v n}}_{t=0}$ vanish unless $n_+=n_-=0$ (see
Appendix \ref{sec:initial_conditions}), so there is a substantially
smaller number of terms with non-zero contribution to the correlator
$\bk{\S_{\v m}}$, and therefore a substantially smaller number of
non-zero terms that are neglected in the truncated Taylor expansion.

Although the truncated Taylor expansion breaks down at short times, it
has two key advantages over other methods to compute collective spin
correlators.  First, computing spin correlators with the truncated
Taylor expansion is generally much faster and requires much fewer
computing resources than alternatives.  The quantum trajectory Monte
Carlo simulations performed for Figure \ref{fig:squeezing_incoherent},
for example, take $\sim10^4$ CPU hours to compute on standard modern
computing hardware; the bulk of this time is spent performing sparse
matrix multiplication, leaving little room to further optimize run
time.  Parallelization can reduce the actual run time of these quantum
trajectory simulations to $\sim10$ hours, but at the cost of greatly
increasing computing resource requirements.  The truncated Taylor
expansion results in Figure \ref{fig:squeezing_incoherent}, meanwhile,
take $\sim10$ seconds to compute with a single CPU on similar
computing hardware.

@@

These are very easy to compute.

@@

The one-axis twisting (OAT) Hamiltonian for $N$ spin-1/2 particles
takes the form
\begin{align}
  H_{\t{OAT}}
  = \chi S_\z^2
  = \f12 \chi \sum_{j<k} \sigma_\z^{(j)} \sigma_\z^{(k)} + \f14 N \chi,
\end{align}
where $\sigma_\z^{(j)}$ represents a Pauli-$z$ operator acting on spin
$j$.  This model is a special case of the zero-field Ising Hamiltonian
previously solved in Ref.~[\citenum{foss-feig2013nonequilibrium}] via
exact, analytical treatment of the quantum trajectory Monte Carlo
method for computing expectation values.  The solution therein
accounts for coherent evolution in addition to decoherence via
uncorrelated single-spin decay, excitation, and dephasing respectively
at rates $\gamma_-$, $\gamma_+$, and $\gamma_\z$ (denoted by
$\Gamma_{\t{ud}}$, $\Gamma_{\t{du}}$, and $\Gamma_{\t{el}}$ in
Ref.~[\citenum{foss-feig2013nonequilibrium}]).  Letting $S\equiv N/2$
and $\mu,\nu\in\set{+1,-1}$, we adapt expectation values computed in
Ref.~[\citenum{foss-feig2013nonequilibrium}] for the initial state
$\ket{\X}$ satisfying $S_\x\ket{\X}=S\ket{\X}$ evolving under
$H_{\t{OAT}}$, finding
\begin{align}
  \bk{S_+}
  &= S e^{-\Gamma t} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_\z}
  &= -\f{\mu}{2}\bk{S_\mu} + S \p{S-\f12} e^{-\Gamma t}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \delta_{\mu,-\nu} \p{S + \mu\bk{S_\z}}
  + S \p{S-\f12} e^{-2\Gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{X,t}
  &\equiv e^{-\lambda t} \sp{\cos\p{t\sqrt{s_X^2-r}}
    + \lambda t~\t{sinc}\p{t\sqrt{s_X^2-r}}},
  \\
  \Psi\p{X,t}
  &\equiv e^{-\lambda t} \p{is_X-\gamma}t~
  \t{sinc}\p{t\sqrt{s_X^2-r}},
\end{align}
for
\begin{align}
  \gamma \equiv -\f12 \p{\gamma_+ - \gamma_-},
  &&
  \lambda \equiv \f12 \p{\gamma_+ + \gamma_-},
  &&
  r \equiv \gamma_+ \gamma_-,
  &&
  \Gamma \equiv \f12\gamma_\z + \lambda,
  &&
  s_X \equiv X + i\gamma,
\end{align}
In order to compute spin squeezing as measured by $\xi^2$ in
\eqref{eq:squeezing}, we additionally need analytical expressions for
$\bk{S_\z}$ and $\bk{S_\z^2}$.  As these operators commute with both
the OAT Hamiltonian and the single-spin operators $\sigma_\z^{(j)}$,
their evolution is governed entirely by decay-type decoherence (see
Appendix \ref{sec:decay_single}), which means
\begin{align}
  \f{d}{dt} S_\z
  &= S\p{\gamma_+-\gamma_-} - \p{\gamma_++\gamma_-} S_\z,
  \\
  \f{d}{dt}\p{S_\z^2}
  &= S\p{\gamma_++\gamma_-} + 2\p{\gamma_+-\gamma_-}\p{S-\f12} S_\z
  - 2 \p{\gamma_++\gamma_-} S_\z^2.
\end{align}
The initial conditions $\bk{S_\z}_{t=0}=0$ and $\bk{S_\z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_\z}
  &= S\p{\f{\gamma_+-\gamma_-}{\gamma_++\gamma_-}}
  \p{1-e^{-\p{\gamma_+ + \gamma_-} t}}, \\
  \bk{S_\z^2} &= S \sp{\f12 + \p{S-\f12} \f{\bk{S_\z}^2}{S^2}}.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}-\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the spin squeezing
parameter $\xi^2$ in \eqref{eq:squeezing} at any time throughout
evolution of the initial state $\ket{\X}$ under $H_{\t{OAT}}$, under
appropriate assumptions about the relevant sources of decoherence.

@@

In principle, the expressions in \eqref{eq:D_mn_k} and
\eqref{eq:C_mn_E}-\eqref{eq:E_mn_p} can be used to evaluate the
commutator $\sp{S_\mu^m,S_\bmu^n}_- = C_{mn;\mu}$, but these result
are -- put lightly -- quite a mess: the expression for
$E_{mn;\mu}^{(p)}$ in \eqref{eq:E_mn_p} involves a sum over $p$
mutually dependent intermediate variables, each term of which
additionally contains a product of $p$ factors.  We therefore devote
the rest of this section to simplifying our result for the commutator

@@

>>>>

In this work, we present a new method to simulate short-time dynamics
of collective spin systems with permutationally-symmetric decoherence.
This method is based on a short-time expansion of exact solutions to
the equations of motion for Heisenberg operators.  Evaluating this
expansion requires using the structure constants of a collective spin
operator algebra; the calculation of these structure constants (in
Appendices \ref{sec:identities}--\ref{sec:prod_general}) is one of the
main technical results of this work, which we hope will empower both
analytical and numerical future studies of collective spin systems.
We benchmark our method against both exact and quantum trajectory
Monte Carlo computations of spin squeezing in accessible parameter
regimes, highlighting both advantages and limitations of our
short-time expansion.  Finally, we showcase an application of our
method with computations that are inaccessible to other numerical
methods.

@@

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$s_{n=\x,\y,\z}\equiv\sigma_n/2$ and
$s_\pm\equiv s_\x\pm is_\y=\sigma_\pm$ with Pauli operators
$\sigma_n$, we denote an operator that acts with $s_n$ on the spin
indexed by $j$ and trivially (i.e.~with the identity $\1$) on all
other spins by $s_n^{(j)}$.  We then define the collective spin
operators $S_n\equiv\sum_{j=1}^Ns_n^{(j)}$ for
$n\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\S_{\v m}}$ as a
basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$, we can expand all
collective spin Hamiltonians in the form
\begin{align}
  H = \sum_{\v m} h_{\v m} \S_{\v m},
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{\v m}^*=h_{\v m^*}$ for
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The time evolution of a particular
Heisenberg operator $\S_{\v m}$ under a Hamiltonian of the form in
Eq.~\eqref{eq:H_general} is then given by
\begin{align}
  \f{d}{dt}\S_{\v m}
  &= i\sum_{\v n}h_{\v n}\sp{\S_{\v n},\S_{\v m}}_-
  + \sum_\J \gamma_\J \D\p{\J;\S_{\v m}} \\
  &= \sum_{\v n} T_{\v m\v n} \S_{\v n},
  \label{eq:EOM}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\J$ is a set of jump operators
with a corresponding decoherence rate $\gamma_\J$;
\begin{align}
  \D\p{\J;\O}
  \equiv \sum_{J\in\J} \p{J^\dag \O J - \f12\sp{J^\dag J,\O}_+}
\end{align}
is the Heisenberg picture dissipator, or Lindblad superoperator, for
all jump operators $J\in\J$; and $T_{\v m\v n}$ is a matrix element of
the time derivative operator.  Decoherence via uncorrelated decay of
individual spins at a rate $\gamma_-$, for example, would be described
by the set of jump operators $\J_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$
with $\gamma_{\J_-}=\gamma_-$.  The commutator in Eq.~\eqref{eq:EOM}
can be determined from the structure constants $f_{\v\ell\v m\v n}$ of
the general product
$\S_{\v\ell}\S_{\v m}=\sum_{\v n}f_{\v\ell\v m\v n}\S_{\v n}$ worked
out in Appendices \ref{sec:identities}--\ref{sec:prod_general}, while
the effects of decoherence from jump operators (i.e.~elements of $\J$)
of the form $\gamma^{(j)}=\sum_n\gamma_ns_n^{(j)}$ and
$\Gamma=\sum_n\Gamma_nS_n$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be the main technical contributions of
our work, with potential applications beyond the short-time simulation
method presented here.  These ingredients are sufficient to compute
matrix elements $T_{\v m\v n}$ of the time derivative operator $d/dt$
in Eq.~\eqref{eq:EOM} in most cases of practical interest.

@@

We can expand Heisenberg operators in a Taylor series about time $t=0$
to write
\begin{align}
  \S_{\v m}\p{t}
  % = \left. \exp\p{t \f{d}{d\tau}} \S_{\v m}\p{\tau} \right|_{\tau=0}
  &= \sum_{k\ge0} \f{t^k}{k!}
  \left. \f{d^k}{d\tau^k} \S_{\v m}\p{\tau} \right|_{\tau=0} \\
  &= \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \S_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_1} T_{\v p_1\v p_2} T_{\v p_2\v p_3}
  \cdots T_{\v p_{k-1}\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \S_{\v m}\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v n} T^k_{\v m\v n} \S_{\v n}\p{0}.
  \label{eq:order_truncation}
\end{align}
Unlike the truncation in Eq.~\eqref{eq:weight_truncation}, the nonzero
matrix elements $T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ now tell us
which operators are relevant for computing the expectation value
$\bk{\S_{\v m}\p{t}}$ to a fixed order $M$.  Note that the relation
$\S_{\v m^*}=\S_{\v m}^\dag$, which by Hermitian conjugation of
Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can be exploited to cut both the
number of initial-time expectation values $\bk{\S_{\v m}\p{0}}$ and
the number of matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the Taylor expansion in Eq.~\eqref{eq:time_series} implies for any
given time $t$, there exists a truncation order $M_t$ for which the
truncation error in Eq.~\eqref{eq:order_truncation} is negligibly
small.  In practice, only a maximal truncation order $M_{\t{max}}$ is
accessible with limited computational resources, such that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} only
allows computation of the correlator $\bk{\S_{\v m}\p{t}}$ to a
maximal time $t_{M_{\t{max}}}$.  As we will see in the following
section, $M_{\t{max}}=35$ will suffice for the computation of
collective spin correlators up to times that are relevant e.g.~for
spin squeezing protocols.

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is the
directionally-unbiased Ramsey squeezing parameter determined by the
maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$ over
that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\v S\equiv\p{S_\x,S_\y,S_\z}$ is a collective spin vector, and
the minimization is performed over all unit vectors $\uv n$ in the
plane orthogonal to the mean spin vector $\bk{\v S}$.  This squeezing
parameter is entirely determined by one- and two-spin correlators of
the form $\bk{S_m}$ and $\bk{S_m S_n}$.  In the case of the coherent
dynamics discussed here, these correlators are obtainable via exact
simulations of quantum dynamics in the $\p{N+1}$-dimensional Dicke
manifold of states $\set{\ket{S,m}}$ with net spin $S$ and spin
projection $m$ onto the $z$ axis, i.e.~with
$\bk{S,m|\v S^2|S,m}=S\p{S+1}$ and $\bk{S,m|S_\z|S,m}=m$ for
$m\in\set{-S,-S+1,\cdots,S}$.  In the presence of
permutationally-symmetric decoherence, these correlators are
obtainable with the collective-spin quantum trajectory Monte Carlo
method developed in ref.~\cite{zhang2018montecarlo}.  In this work,
these exact and quantum trajectory simulations will be used to
benchmark the truncated Taylor expansion in
Eq.~\eqref{eq:order_truncation}.

@@

The main lesson from Figure \ref{fig:benchmarking} is that the
truncated Taylor expansion in Eq.~\eqref{eq:order_truncation} gives
essentially exact results right up until a sudden and drastic
departure that is simple to diagnose by inspection.  This drastic
departure occurs because the truncated Taylor expansion neglects
high-weight (i.e.~large-$\abs{\v n}$) operators $\S_{\v n}$ whose
contributions to a correlator of interest $\bk{\S_{\v m}}$ eventually
become non-negligible.  For sufficiently large times $t$, these
individual contributions may generally be large compared to the value
of $\bk{\S_{\v m}}$, in which case their truncation will yield
non-physical results.  This breakdown mechanism is also the reason for
prolonged agreement between the truncated Taylor expansion and
benchmarking methods when squeezing the initial state $\ket{-\Z}$
rather than $\ket{\X}$: for initial states $\ket{-\Z}$, all
initial-time correlators $\bk{\S_{\v n}\p{0}}$ vanish unless
$n_+=n_-=0$ (see Appendix \ref{sec:initial_conditions}), so there is a
substantially smaller number of terms with non-zero contribution to
the correlator $\bk{\S_{\v m}}$, and therefore a substantially smaller
number of non-zero terms that are neglected in the truncated Taylor
expansion.

Although the truncated Taylor expansion breaks down at short times, it
has two key advantages over other methods to compute collective spin
correlators.  First, computing spin correlators with the truncated
Taylor expansion is generally much faster and requires much fewer
computing resources than the alternatives.  The quantum trajectory
Monte Carlo simulations performed for Figure
\ref{fig:squeezing_incoherent}, for example, take $\sim10^4$ CPU hours
to compute on standard modern computing hardware; the bulk of this
time is spent performing sparse matrix multiplication, leaving little
room to further optimize run time.  Parallelization can reduce the
actual run time of these quantum trajectory simulations to $\sim10$
hours, but at the cost of greatly increasing computing resource
requirements.  The truncated Taylor expansion results in Figure
\ref{fig:squeezing_incoherent}, meanwhile, take $\sim10$ seconds to
compute with a single CPU on similar computing hardware.

@@

As a final example, we consider the calculation of two-time
correlation functions in collective spin systems.

@@

The one-axis twisting (OAT) Hamiltonian for $N$ spin-1/2 particles
takes the form
\begin{align}
  H_{\t{OAT}}
  = \chi S_\z^2
  = \f12 \chi \sum_{j<k} \sigma_\z^{(j)} \sigma_\z^{(k)} + \f14 N \chi,
\end{align}
where $\sigma_\z^{(j)}$ represents a Pauli-$z$ operator acting on spin
$j$.  This model is a special case of the zero-field Ising Hamiltonian
previously solved in Ref.~[\citenum{foss-feig2013nonequilibrium}] via
exact, analytical treatment of the quantum trajectory Monte Carlo
method for computing expectation values.  The solution therein
accounts for coherent evolution in addition to decoherence via
uncorrelated single-spin decay, excitation, and dephasing respectively
at rates $\gamma_-$, $\gamma_+$, and $\gamma_\z$ (denoted by
$\Gamma_{\t{ud}}$, $\Gamma_{\t{du}}$, and $\Gamma_{\t{el}}$ in
Ref.~[\citenum{foss-feig2013nonequilibrium}]).  Letting $S\equiv N/2$
and $\mu,\nu\in\set{+1,-1}$, we adapt expectation values computed in
Ref.~[\citenum{foss-feig2013nonequilibrium}] for the initial state
$\ket{\X}$ satisfying $S_\x\ket{\X}=S\ket{\X}$ evolving under
$H_{\t{OAT}}$, finding
\begin{align}
  \bk{S_+}
  &= S e^{-\Gamma t} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_\z}
  &= -\f{\mu}{2}\bk{S_\mu} + S \p{S-\f12} e^{-\Gamma t}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \delta_{\mu,-\nu} \p{S + \mu\bk{S_\z}}
  + S \p{S-\f12} e^{-2\Gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{X,t}
  &\equiv e^{-\lambda t} \sp{\cos\p{t\sqrt{s_X^2-r}}
    + \lambda t~\t{sinc}\p{t\sqrt{s_X^2-r}}},
  \\
  \Psi\p{X,t}
  &\equiv e^{-\lambda t} \p{is_X-\gamma}t~
  \t{sinc}\p{t\sqrt{s_X^2-r}},
\end{align}
for
\begin{align}
  \gamma \equiv -\f12 \p{\gamma_+ - \gamma_-},
  &&
  \lambda \equiv \f12 \p{\gamma_+ + \gamma_-},
  &&
  r \equiv \gamma_+ \gamma_-,
  &&
  \Gamma \equiv \f12\gamma_\z + \lambda,
  &&
  s_X \equiv X + i\gamma,
\end{align}
In order to compute spin squeezing as measured by $\xi^2$ in
\eqref{eq:squeezing}, we additionally need analytical expressions for
$\bk{S_\z}$ and $\bk{S_\z^2}$.  As these operators commute with both
the OAT Hamiltonian and the single-spin operators $\sigma_\z^{(j)}$,
their evolution is governed entirely by decay-type decoherence (see
Appendix \ref{sec:decay_single}), which means
\begin{align}
  \f{d}{dt} S_\z
  &= S\p{\gamma_+-\gamma_-} - \p{\gamma_++\gamma_-} S_\z,
  \\
  \f{d}{dt}\p{S_\z^2}
  &= S\p{\gamma_++\gamma_-} + 2\p{\gamma_+-\gamma_-}\p{S-\f12} S_\z
  - 2 \p{\gamma_++\gamma_-} S_\z^2.
\end{align}
The initial conditions $\bk{S_\z}_{t=0}=0$ and $\bk{S_\z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_\z}
  &= S\p{\f{\gamma_+-\gamma_-}{\gamma_++\gamma_-}}
  \p{1-e^{-\p{\gamma_+ + \gamma_-} t}}, \\
  \bk{S_\z^2} &= S \sp{\f12 + \p{S-\f12} \f{\bk{S_\z}^2}{S^2}}.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}--\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the spin squeezing
parameter $\xi^2$ in \eqref{eq:squeezing} at any time throughout
evolution of the initial state $\ket{\X}$ under $H_{\t{OAT}}$, under
appropriate assumptions about the relevant sources of decoherence.

@@

In principle, the expressions in \eqref{eq:D_mn}, \eqref{eq:D_mn_k},
\eqref{eq:C_mn_E}, and \eqref{eq:E_mn_p} suffice to evaluate the
commutator $\sp{S_\mu^m,S_\bmu^n}_- = C_{mn;\mu}$, but this result is
-- put lightly -- quite a mess: the expression for $E_{mn;\mu}^{(p)}$
in \eqref{eq:E_mn_p} involves a sum over $p$ mutually dependent
intermediate variables, each term of which additionally contains a
product of $p$ factors.  We therefore devote the rest of this section
to simplifying our result for the commutator

@@

  \sum_{\ell=0}^p \epsilon_{mn}^{p\ell} \p{\mu S_\z}^\ell.
